---
title: "P8105_HW2"
date: "2022-10-04"
output: html_document
---

```{r initial, message = FALSE}
library(tidyverse)
library(readxl)
```

###    Problem 1

Below we import and clean data from `NYC_Transit_Subway_Entrance_And_Exit_Data.csv`. The process begins with data import, updates variable names, and selects the columns that will be used in later parts fo this problem. We update `entry` from `yes` / `no` to a logical variable. As part of data import, we specify that `Route` columns 8-11 should be character for consistency with 1-7.

```{r}
trans_ent = 
  read_csv(
    "~/Documents/Columbia Grad School/Fall 2022/Data Science/HW/p8105_hw2_mmu2108/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>% 
  janitor::clean_names() %>% 
  select(
    line, station_name, station_latitude, station_longitude, 
    starts_with("route"), entry, exit_only, vending, entrance_type, 
    ada) %>% 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))
```

As it stands, these data are not "tidy": route number should be a variable, as should route. That is, to obtain a tidy dataset we would need to convert `route` variables from wide to long format. This will be useful when focusing on specific routes, but may not be necessary when considering questions that focus on station-level variables. 

The following code chunk selects station name and line, and then uses `distinct()` to obtain all unique combinations. As a result, the number of rows in this dataset is the number of unique stations.

```{r}
trans_ent %>% 
  select(station_name, line) %>% 
  distinct
```

The next code chunk is similar, but filters according to ADA compliance as an initial step. This produces a dataframe in which the number of rows is the number of ADA compliant stations. 

```{r}
trans_ent %>% 
  filter(ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

To compute the proportion of station entrances / exits without vending allow entrance, we first exclude station entrances that do not allow vending. Then, we focus on the `entry` variable -- this logical, so taking the mean will produce the desired proportion (recall that R will coerce logical to numeric in cases like this).

```{r}
trans_ent %>% 
  filter(vending == "NO") %>% 
  pull(entry) %>% 
  mean
```

Lastly, we write a code chunk to identify stations that serve the A train, and to assess how many of these are ADA compliant. As a first step, we tidy the data as alluded to previously; that is, we convert `route` from wide to long format. After this step, we can use tools from previous parts of the question (filtering to focus on the A train, and on ADA compliance; selecting and using `distinct` to obtain dataframes with the required stations in rows).

```{r}
trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A") %>% 
  select(station_name, line) %>% 
  distinct

trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A", ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

### Problem 2
```{r mr_tw}
mr_trash = read_excel("~/Documents/Columbia Grad School/Fall 2022/Data Science/HW/p8105_hw2_mmu2108/trash_wheel_collection_data.xlsx", sheet = "Mr. Trash Wheel") %>% 
  janitor::clean_names() %>%
  drop_na(, dumpster) %>%
  select(-x15, -x16) %>%
  mutate(, year = as.integer(year)) %>%
  mutate(, sports_balls = round(sports_balls, digits = 0),
         sports_balls = as.integer(sports_balls)) %>%
  mutate(vessel = 'mr')
```

```{r prof_tw}
prof_trash = read_excel("~/Documents/Columbia Grad School/Fall 2022/Data Science/HW/p8105_hw2_mmu2108/trash_wheel_collection_data.xlsx", sheet = "Professor Trash Wheel") %>% 
  janitor::clean_names() %>%
  drop_na(, dumpster) %>%
  mutate(vessel = 'professor')
```


```{r trash_merge}
trashwheel = 
  bind_rows(mr_trash, prof_trash) %>%
  janitor::clean_names() %>%
  select(vessel, everything()) 
```

In total, there are `r count(trashwheel)` dumpster observations among Mr. Trash Wheel and Professor Trash Wheel.  It is notable that Professor Trash Wheel did not collect any sports balls, thus all observations are marked as 'NA' in the combined 'trashwheel' dataset.

Professor Trash Wheel collected a total of `r sum(prof_trash$weight_tons)` tons of trash.  There was a total of `r sum(mr_trash$sports_balls, na.rm = TRUE)` sports balls collected by Mr. Trash Wheel.

### Problem 3
```{r pols_clean}
pols_tidy = 
  read_csv(
    "~/Documents/Columbia Grad School/Fall 2022/Data Science/HW/p8105_hw2_mmu2108/fivethirtyeight_datasets/pols-month.csv") %>%
  janitor::clean_names() %>%
  separate(mon, into = c("year", "month", "day"), sep = "-") %>%
  mutate(month = as.numeric(month)) %>%
  arrange(year, month) %>%
  mutate(month = month.name[month]) %>%
  mutate(year = as.numeric(year)) %>%
  unite(president, prez_gop, prez_dem, remove = TRUE) %>%
  select(-day)
```

```{r snp_clean}
snp_tidy = 
  read_csv(
    "~/Documents/Columbia Grad School/Fall 2022/Data Science/HW/p8105_hw2_mmu2108/fivethirtyeight_datasets/snp.csv") %>%
  janitor::clean_names() %>%
  separate(date, into = c("month", "day", "year"), sep = "/") %>%
  mutate(year = as.integer(year)) %>%
  mutate(year = ifelse(year <= 15, year + 2000, year + 1900)) %>%
  mutate(month = as.numeric(month)) %>%
  arrange(year, month) %>%
  mutate(month = month.name[month]) %>%
  select(year, month, everything()) %>%
  mutate(year = as.numeric(year)) %>%
  select(-day)
```

```{r unemploy_clean}
unemploy_tidy = 
  read_csv(
    "~/Documents/Columbia Grad School/Fall 2022/Data Science/HW/p8105_hw2_mmu2108/fivethirtyeight_datasets/unemployment.csv") %>%
  janitor::clean_names() %>%
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment"
  ) %>%
  mutate(month = match(month, tolower(month.abb))) %>%
  mutate(month = month.name[month]) %>%
  mutate(year = as.numeric(year))
```

```{r}
fivethirtyeight_merge = 
  merge(pols_tidy, snp_tidy, by = c("year", "month"))
  
fivethirtyeight_merge = 
  merge(fivethirtyeight_merge, unemploy_tidy, by = c("year", "month"))
```


